Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwest China using organic residue analysis

### 2023-03-15 analysis/paper/paper.Rmd
corrected some references

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015; @Weisskopf2015]. The Neolithic Hengduan Mountain regions are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as those grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Lin2022], and millet could have already become part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

Although the recent findings at Guijiabao and Gaoshan support mixed agriculture, it is still unclear how much millet or rice was included in local culinary practices during the Neolithic period. The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Hao2022]. This indicates there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis for ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

<!--discussion about possible ruminant fat origin-->
The ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer, sheep, or horses fed on a C3 diet based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017]. This also corresponds with unearthed faunal remains from the sites and the local natural environments. Stone artifacts found at the site, including stone balls and arrowheads, further indicates hunting activities [@Hao2022; @Guijiabao2020a]. However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China, with only a sporadic presence in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on. Thus, supported by faunal evidence, deer is the most likely part of the local food choices. It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these species have been found in southwest China [@Zhang2016; @Chen2020].

>>>>

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas usually accompany the exchange of goods, techniques, or ideas that could have cultural or economic impacts on those groups involved in the interaction [@Burmeister2000]. In southwestern China, the Hengduan Mountain ranges and valleys connecting the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau play an important role in the cultural exchange and spread of crops into this region and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate between rice/millet dispersal waves from northwest China and the middle Yangtze valley [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwest Sichuan around 4,700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Neolithic Hengduan Mountain regions are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as those grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether the new crops or species became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues from archaeological vessels at Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan so far excavated [@Guijiabao2017a; @Guijiabao2020a]. The recent evidence at Guijiabao reveals that millet and rice were both present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016], and millet could have already become part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming, we still lack direct evidence of culinary practices during this period. To understand the extent of new crops/species processing at Guijiabao, we ask the questions: What were the culinary practices reflected by pottery use? Could we identify the evidence of new crops or species incorporated into the local food practices during the Neolithic period? We answer those questions by examining potsherds using organic residue analysis, which allows us to directly explore food processed in ceramic vessels.

@@

Although the recent findings at Guijiabao and Gaoshan support mixed agriculture, it is still unclear how much millet or rice was included in local culinary practices during the Neolithic period. The macro-botanical and phytolith analyses for Guijiabao show that millet is the major crop, similar to the findings at the Yingpanshan site in Sichuan [@Zhao2011]. However, plant remains from many sites in southwest China reveal a predominance of rice, such as Gaoshan and Baodun in Sichuan [@JiangGaoshan2017] and Henglanshan and Baiyangcun in Yunnan [@Jiang2016; @Henglanshan2014]. This indicates there were variations in agricultural practices in southwest China. To better understand the extent of the adoption of new crops, organic residue analysis for ceramics is a useful approach since we can directly study food residues related to culinary practices, including food preparation, processing, and cooking. This provides another way to examine the expansion of agriculture in the Yanyuan area.

@@

<!--discussion about possible ruminant fat origin-->
The ruminant fats identified in the Neolithic potsherds (ca. 4500 BP) from Guijiabao and Gaoshan could be derived from deer, sheep, or horses fed on a C3 diet based on their isotopic values that are consistent with global references [@Evershed2002; @Mileto2017]. This finding, especially the consumption of deer, also corresponds with unearthed faunal remains from the sites and the local natural environments. Stone artifacts found at the site, including stone balls and arrowheads, further indicates hunting activities [@Hao2022; @Guijiabao2020a]. However, whether horse was part of local food choices in southwest China around 4500 years ago is skeptical. First, the Pleistocene and Neolithic sites with wild horse bones are primarily concentrated in northern China, with only a sporadic presence in the south [@Yuan2006]. Second, domesticated horses first found in southwest China date to the Dian culture (4th century BC to 1st century AD) in eastern Yunnan [@Lei2009; @Chiou2004], which is 3000 years later than the Neolithic period we focus on. Thus, supported by faunal evidence, deer is the most likely part of the local food choices. It is also possible that local wild bovids, such as yaks or water buffalo, were food sources during this period, as these species have been found in sites of southwest China [@Zhang2016; @Chen2020].
